---
title: "R Notebook"
output: html_notebook
---

## Quiz 3


```{r load}
library(SummarizedExperiment)
library(Biobase)
library(GenomicRanges)
library(broom)
library(limma)
library(dplyr)
library(snpStats)
library(tidyr)
```

Question 1

Related Documentation

* http://jtleek.com/genstats/inst/doc/03_04_GLMs.html
* https://bioconductor.org/packages/release/bioc/vignettes/snpStats/inst/doc/data-input-vignette.pdf

Load the example SNP data with the following code:

```{r}
data(for.exercise)
use <- seq(1, ncol(snps.10), 10)
sub.10 <- snps.10[,use]
snpdata = sub.10@.Data
status = subject.support$cc
```


Fit a linear model and a logistic regression model to the data for the 3rd SNP. What are the coefficients for the SNP variable? How are they interpreted? (Hint: Don't forget to recode the 0 values to NA for the SNP data)

```{r q1_ml}
snp3 = as.numeric(snpdata[,3])
snp3[snp3==0] = NA
lm(status ~ snp3)  
```

```{r q1_log}
glm1 = glm(status ~ snp3,family="binomial")
tidy(glm1)
```


Question 6

Load the Montgomery and Pickrell eSet:


```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```

Do the log2(data + 1) transform and fit calculate F-statistics for the difference between studies/populations using genefilter:rowFtests and using genefilter:rowttests. Do you get the same statistic? Do you get the same p-value?